Sparse spectral methods for partial differential equations on spherical caps and bands

### 2020-10-08 papers/spherical-caps/spherical-caps.tex
edits to sc paper

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps and bands. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and Biharmonic operators, showing spectral convergence. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on certain subsets of the sphere: spherical bands and spherical caps. More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: ||(x,y)|| := \sqrt{x^2 + y^2} =  \rho(z)\}
\end{align*}
and consider the case where $\alpha \in (-1,1)$ and where $ \rho(z) := \sqrt{1-z^2}$.
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
For simplicity we focus on a spherical cap with   $\beta := 1$, though there is a simple extension to a spherical band by taking $\beta \in (\alpha,1)$.

We advocate using a basis that is polynomial in cartesian coordinates, that is, polynomial in $x$, $y$, and $z$, and orthogonal with respect to a prescribed weight: that is, multivariate orthogonal polynomials, whose construction was considered in \cite{olver2020orthogonal}. Equivalently, we can think of these as polynomials modulo the vanishing ideal $\{ x^2 + y^2 + z^2 = 1 \}$, or simply as a linear recombination of spherical harmonics that are orthogonalised on a subset of the sphere.    This is in contrast to more standard approaches based on mapping the geometry to a simpler one (e.g., a rectangle or disk) and using orthogonal polynomials in the mapped coordinates (e.g., a basis that is polynomial in the spherical coordinates $\varphi$ and $\theta$). The benefit of the new approach is that we do not need to resolve Jacobians, and thereby we can achieve sparse discretisations for partial differential operators, including those with polynomial variable coefficients.   On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms})
\end{align*}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

@@

 The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical \cite[\S5]{magnus1995painleve}) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations. Furthermore, we can use this framework to derive sparse $p$-finite element methods, which we introduce as an Appendix here. The methods are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

Note that we consider partial differential operators that are not necessarily rotational invariant:  for example, one can use these techniques for  Schr\"odinger operators $\DeltaS + V(x,y,z)$ where $V$ is first approximated by a polynomial. A nice feature though is that if the partial differential operator is invariant with respect to rotation around the $z$ axes (e.g., a Schr\"odinger operator with potential $V(z)$) the discretisation decouples, and can be reordered as a block-diagonal matrix. This improves the complexity further to an optimal $O(p^2$). \sotodoinline{Adding a complexity plot showing this would be great} \bstodoinline{I am not sure how to do a complexity plot}

@@

\noindent\appref{PFEM}: We outline how our framework can allow us to construct sparse $hp$-finite element method on the sphere, using spherical band and cap elements that would capture spherical geometry precisely. We lay the groundwork by simply presenting a sparse $p$-finite element method using a single spherical cap element.

@@

For $J_x^{(a)}$, the sub-blocks $A^{(a)}_{x, k}, C^{(a)}_{x, k}$ have sub-block-bandwidths $(4,4)$ while the sub-blocks $B^{(a)}_{x,k}$ are zero matrices:
\begin{align*}
	B^{(a)}_{x,0} &= 0 \in \R^{(N+1)\times2N}, \quad B^{(a)}_{x,k} = 0 \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \\
	A^{(a)}_{x,k} &:= 
		\begin{pmatrix}
			A^{(a)}_{k,k,6} & & & \\
			A^{(a)}_{k+1,k,4} & \ddots & & \\
			A^{(a)}_{k+2,k,2} & \ddots & \ddots & \\
			& \ddots & \ddots & A^{(a)}_{N-1,k,6} \\
			& & A^{(a)}_{N,k,2} & A^{(a)}_{N,k,4}
		\end{pmatrix}, \quad k = 0,1,2,\dots,N \\
	A^{(a)}_{x,0} &\in \R^{(N+1)\times2N}, \quad A^{(a)}_{x,k} \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \\
	C^{(a)}_{x,k} &:= 
		\begin{pmatrix}
			A^{(a)}_{k,k,1} & A^{(a)}_{k,k,3} & A^{(a)}_{k,k,5} & & \\
			& \ddots & \ddots & \ddots \\
			& & \ddots & \ddots & A^{(a)}_{N-1,k,5} \\
			& & & A^{(a)}_{N,k,1} & A^{(a)}_{N,k,3}
		\end{pmatrix} \in \R^{2(N-k+1)\times2(N-k+2)}, \quad k = 1,2,\dots \\
	C^{(a)}_{x,1} &\in \R^{2N\times(N+1)}, \quad C^{(a)}_{x,k} \in \R^{2(N-k+1)\times2(N-k+3)} \text{ for } k = 2,3,\dots,N
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align}
	A^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			\alphaa_{n,k,j} & 0 \\
			0 & \alphaa_{n,k,j}
		\end{pmatrix} \in \R^{2\times2}, (k \ne 1 \text { for } j \text{ odd}) \label{eqn:jacobisubblocksA1} \\
	A^{(a)}_{n,0,j} &:=
		\begin{pmatrix}
			\alphaa_{n,0,j} & 0
		\end{pmatrix} \in \R^{1\times2}, j \text{ odd} \label{eqn:jacobisubblocksA2} \\
	A^{(a)}_{n,1,j} &:=
		\begin{pmatrix}
			\alphaa_{n,1,j} \\
			0
		\end{pmatrix} \in \R^{2\times1}, j \text{ odd} \label{eqn:jacobisubblocksA3} \\
\end{align}

For $J_y^{(a)}$, the sub-blocks have sub-block-bandwidths $(5,5)$ while the sub-blocks $B^{(a)}_{y,k}$ are zero matrices:
\begin{align*}
	B^{(a)}_{y,0} &= 0 \in \R^{(N+1)\times2N}, \quad B^{(a)}_{y,k} = 0 \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \\
	A^{(a)}_{y,k} &:= 
		\begin{pmatrix}
			B^{(a)}_{k,k,6} & & & \\
			B^{(a)}_{k+1,k,4} & \ddots & & \\
			B^{(a)}_{k+2,k,2} & \ddots & \ddots & \\
			& \ddots & \ddots & B^{(a)}_{N-1,k,6} \\
			& & B^{(a)}_{N,k,2} & B^{(a)}_{N,k,4}
		\end{pmatrix}, \quad k = 0,1,2,\dots,N \\
	A^{(a)}_{y,0} &\in \R^{(N+1)\times2N}, \quad A^{(a)}_{y,k} \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \\
	C^{(a)}_{y,k} &:= 
		\begin{pmatrix}
			B^{(a)}_{k,k,1} & B^{(a)}_{k,k,3} & B^{(a)}_{k,k,5} & & \\
			& \ddots & \ddots & \ddots \\
			& & \ddots & \ddots & B^{(a)}_{N-1,k,5} \\
			& & & B^{(a)}_{N,k,1} & B^{(a)}_{N,k,3}
		\end{pmatrix} \in \R^{2(N-k+1)\times2(N-k+2)}, \quad k = 1,2,\dots \\
	C^{(a)}_{y,1} &\in \R^{2N\times(N+1)}, \quad C^{(a)}_{y,k} \in \R^{2(N-k+1)\times2(N-k+3)} \text{ for } k = 2,3,\dots,N
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align}
	B^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			0 & \betaa_{n,k,0,j} \\
			\betaa_{n,k,1,j} & 0
		\end{pmatrix} \in \R^{2\times2}, (k \ne 1 \text { for } j \text{ odd}) \label{eqn:jacobisubblocksB1} \\
	B^{(a)}_{n,0,j} &:=
		\begin{pmatrix}
			0 & \betaa_{n,0,0,j}
		\end{pmatrix} \in \R^{1\times2}, j \text{ odd}  \label{eqn:jacobisubblocksB2} \\
	B^{(a)}_{n,1,j} &:=
		\begin{pmatrix}
			0 \\
			\betaa_{n,1,1,j}
		\end{pmatrix}\in \R^{2\times1}, j \text{ odd}  \label{eqn:jacobisubblocksB3} \\
\end{align}

For $J_z^{(a)}$, the sub-blocks $B^{(a)}_{z,k}$ have sub-block-bandwidths $(2,2)$, while the sub-blocks $A^{(a)}_{z,k}, \: C^{(a)}_{z,k}$ are zero matrices:
\begin{align*}
	A^{(a)}_{z,0} &= 0 \in \R^{(N+1)\times2N}, \quad A^{(a)}_{z,k} = 0 \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \\
	C^{(a)}_{z,1} &= 0 \in \R^{2N\times(N+1)}, \quad C^{(a)}_{z,k} = 0 \in \R^{2(N-k+1)\times2(N-k+3)} \text{ for } k = 2,3,\dots,N \\
	B^{(a)}_{z,k} &:= 
		\begin{pmatrix}
			\Gamma^{(a)}_{k,k,2} & \Gamma^{(a)}_{k,k,3} & \\
			\Gamma^{(a)}_{k+1,k,1} & \ddots & \ddots & \\
			& \ddots & \ddots & \Gamma^{(a)}_{N-1,k,3} \\
			& & \Gamma^{(a)}_{N,k,1} & \Gamma^{(a)}_{N,k,2}
		\end{pmatrix} \in \R^{(2n+1)\times(2n+1)}  \quad n = 0,1,2,\dots \\
	B^{(a)}_{z,0} &\in \R^{(N+1)\times(N+1)}, \quad B^{(a)}_{z,k} \in \R^{2(N-k+1)\times2(N-k+1)} \text{ for } k = 1,2,\dots,N
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align}
	\Gamma^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			\gamma_{n,k,j} & 0 \\
			0 & \gamma_{n,k,j}
		\end{pmatrix} \in \R^{2\times2}, \label{eqn:jacobisubblocksGamma1} \\
	\Gamma^{(a)}_{n,0,j} &:= \gammaa_{n,0,j}. \label{eqn:jacobisubblocksGamma2}
\end{align}

@@

In this section, we introduce new notation for a different ordering of the OP vector, in order to demonstrate how we can obtain a recurrence for the OPs in matrix-vector form, and derive the associated \enquote{Clenshaw matrices} -- we define the Clenshaw matrices as the matrices derived from the OP's $x,y,z$-multiplication relationships that form the multidimensional three-term recurrences \bstodoinline{added explanation of clenshaw matrix in this context} While in Section \ref{subsection:jacobimats} the ordering used is convenient for the application of Jacobi and differential operators to the vector of coefficients of a given function's expansion, the ordering we will use in this section is convenient for building the OPs and more importantly obtaining the \enquote{Clenshaw matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings.

@@

\bstodoinline{I have added defns (and sparsity proofs) for two other Laplacian operators. In practice, I have found that the non-weighted one is singular - is this correct/expected? The weighted one and the resulting Biharmonic operators both work (as far as I have tested).}
\begin{definition}\label{def:differentialoperators}
	Let $a$ be a nonnegative parameter, and $\tilde a \ge 2$ be a positive integer. Define the operator matrices $D_\varphi^{(a)}, \: W_\varphi^{(a)}, \: D_\theta, \: \calL^{(a)\to(a+\tilde a)}, \: \calL_W^{(a)\to(a-\tilde a)}, \: \Delta^{(1)}_W$ according to:
\begin{align*}
	\rho \pfpx{\varphi}{f} (x,y,z)&= \bigscop_N^{(a+1)}(x,y,z)^\top \: D_\varphi^{(a)} \: \mathbf{f}, \\
	\rho \ppx{\varphi}[\genjacw^{(a,0)}(z) \: f(x,y,z)] &= \bigW_N^{(a-1)}(x,y)^\top \: W_\varphi^{(a)}\: \mathbf{f}, \\
	\pfpx{\theta}{f}(x,y,z) &= \bigscopNa(x,y,z)^\top \: D_\theta \: \mathbf{f}, \\
	\DeltaS f(x,y,z) &= \bigscopN^{(a+\tilde a)}(x,y,z)^\top \: \calL^{(a)\to(a+\tilde a)} \: \mathbf{f}, \\
	\DeltaS \big( \genjacw^{(a,0)}(z) \: f(x,y,z) \big) &= \bigW_N^{(a-\tilde a)}(x,y,z)^\top \: \calL_W^{(a)\to(a-\tilde a)} \: \mathbf{f},  \quad \text{ (for } a \ge 2 \text{ only)} \\
	\DeltaS \big( \genjacw^{(1,0)}(z) \: f(x,y,z) \big) &= \bigscopN^{(1)}(x,y,z)^\top \: \Delta^{(1)}_W \: \mathbf{f}, \quad \text{ (for } a = 1 \text{ only)}
\end{align*}
\end{definition}

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\begin{align*}
	u(x,y,z) &= 0 \quad \text{for } (x,y,z) \in \Omega, \\
	\pfpx{n}{u}(x,y,z) &= \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
\end{align*}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(\vec{x}) = {\vec{x} \over \norm{\vec{x}}}$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\begin{align*}
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
\end{align*}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions. The sparsity and structure of this biharmonic operator are seen in Figure \ref{fig:sparsity}.

>>>>

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, disk-slices and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps and bands. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and Biharmonic operators, showing spectral convergence. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on certain subsets of the sphere -- specifically spherical caps. More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: ||(x,y)|| := \sqrt{x^2 + y^2} = \rho(z)\}
\end{align*}
and consider the case where $\alpha \in (-1,1)$, $\beta := 1$ and where $ \rho(z) := \sqrt{1-z^2}$.
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
For simplicity we focus on the case of a spherical cap, though there is a simple extension to a spherical band by taking $\beta \in (\alpha,1)$. The methods presented here translate to the spherical band case by including the necessary adjustments to the weights and recurrence relations we present in this paper (these adjustments make the mathematics more involved, which is why they are omitted here, but the approach is the same). For clarity we note that, for all examples and plots in this paper, we take $\alpha = 0.2$. \bstodo{removed "band" from title, and added a bit more here about the extension}

We advocate using a basis that is polynomial in cartesian coordinates, that is, polynomial in $x$, $y$, and $z$, and orthogonal with respect to a prescribed weight: that is, multivariate orthogonal polynomials, whose construction was considered in \cite{olver2020orthogonal}. Equivalently, we can think of these as polynomials modulo the vanishing ideal $\{ x^2 + y^2 + z^2 = 1 \}$, or simply as a linear recombination of spherical harmonics that are orthogonalised on a subset of the sphere. This is in contrast to more standard approaches based on mapping the geometry to a simpler one (e.g., a rectangle or disk) and using orthogonal polynomials in the mapped coordinates (e.g., a basis that is polynomial in the spherical coordinates $\varphi$ and $\theta$). The benefit of the new approach is that we do not need to resolve Jacobians, and thereby we can achieve sparse discretisations for partial differential operators, including those with polynomial variable coefficients. Further, we remove the singular nature as $\alpha$ approaches $0$ that such a projection would give, since our new approach is applicable for all $\alpha \in (-1,1)$. On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms})
\end{align*}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

@@

 The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical \cite[\S5]{magnus1995painleve}) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations.

 \begin{figure}[t]
	\centering % <-- added
	\includegraphics[scale=0.3]{complexity-Nend=1000}
	\caption{Complexity plot showing the length of time taken to construct the operator matrix for $\DeltaS + V(x,y,z)$, with rotationally invariant potential $V(x,y,z) = \cos(z)$.}
	\centering
	\label{fig:complexity}
\end{figure}

Note that we consider partial differential operators that are not necessarily rotational invariant:  for example, one can use these techniques for  Schr\"odinger operators $\DeltaS + V(x,y,z)$ where $V$ is first approximated by a polynomial. A nice feature though is that if the partial differential operator is invariant with respect to rotation around the $z$ axes (e.g., a Schr\"odinger operator with potential $V(z)$) the discretisation decouples, and can be reordered as a block-diagonal matrix. This improves the complexity further to an optimal $O(p^2$), which is demonstrated in Figure \ref{fig:complexity}. This plot presents the time taken (measured using the \enquote{@belapsed} macro from the BenchmarkTools package in Julia) to construct the operator for $\DeltaS + V(x,y,z)$, with $V(x,y,z) = \cos(z)$, as we increase the orthogonal polynomial degree $p$. \bstodoinline{Added plot}

@@

%\noindent\appref{PFEM}: We outline how our framework can allow us to construct sparse $hp$-finite element method on the sphere, using spherical band and cap elements that would capture spherical geometry precisely. We lay the groundwork by simply presenting a sparse $p$-finite element method using a single spherical cap element.

@@

For $J_x^{(a)}$, the sub-blocks $A^{(a)}_{x, k}, C^{(a)}_{x, k}$ have sub-block-bandwidths $(4,4)$ while the sub-blocks $B^{(a)}_{x,k}$ are zero matrices:
\begin{align*}
	B^{(a)}_{x,0} &= 0 \in \R^{(N+1)\times(N+1)}, \quad B^{(a)}_{x,k} = 0 \in \R^{2(N-k+1)\times2(N-k+1)} \text{ for } k = 1,2,\dots,N \\
	A^{(a)}_{x,k} &:= 
		\begin{pmatrix}
			A^{(a)}_{k,k,6} & & & \\
			A^{(a)}_{k+1,k,4} & \ddots & & \\
			A^{(a)}_{k+2,k,2} & \ddots & \ddots & \\
			& \ddots & \ddots & A^{(a)}_{N-1,k,6} \\
			& & A^{(a)}_{N,k,2} & A^{(a)}_{N,k,4}
		\end{pmatrix}, \quad k = 0,1,2,\dots,N \\
	&\big( \text{so that } A^{(a)}_{x,0} \in \R^{(N+1)\times2N}, \quad A^{(a)}_{x,k} \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \big) \\
	C^{(a)}_{x,k} &:= 
		\begin{pmatrix}
			A^{(a)}_{k,k,1} & A^{(a)}_{k,k,3} & A^{(a)}_{k,k,5} & & \\
			& \ddots & \ddots & \ddots \\
			& & \ddots & \ddots & A^{(a)}_{N-1,k,5} \\
			& & & A^{(a)}_{N,k,1} & A^{(a)}_{N,k,3}
		\end{pmatrix}, \quad k = 1,2,\dots,N \\
	&\big( \text{so that } C^{(a)}_{x,1} \in \R^{2N\times(N+1)}, \quad C^{(a)}_{x,k} \in \R^{2(N-k+1)\times2(N-k+2)} \text{ for } k = 2,3,\dots,N \big)
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align}
	A^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			\alphaa_{n,k,j} & 0 \\
			0 & \alphaa_{n,k,j}
		\end{pmatrix} \in \R^{2\times2}, (k \ne 1 \text { for } j \text{ odd}) \label{eqn:jacobisubblocksA1} \\
	A^{(a)}_{n,0,j} &:=
		\begin{pmatrix}
			\alphaa_{n,0,j} & 0
		\end{pmatrix} \in \R^{1\times2}, j \text{ odd} \label{eqn:jacobisubblocksA2} \\
	A^{(a)}_{n,1,j} &:=
		\begin{pmatrix}
			\alphaa_{n,1,j} \\
			0
		\end{pmatrix} \in \R^{2\times1}, j \text{ odd} \label{eqn:jacobisubblocksA3} \\
\end{align}

For $J_y^{(a)}$, the sub-blocks have sub-block-bandwidths $(5,5)$ while the sub-blocks $B^{(a)}_{y,k}$ are zero matrices:
\begin{align*}
	B^{(a)}_{y,0} &= 0 \in \R^{(N+1)\times(N+1)}, \quad B^{(a)}_{y,k} = 0 \in \R^{2(N-k+1)\times2(N-k+1)} \text{ for } k = 1,2,\dots,N \\
	A^{(a)}_{y,k} &:= 
		\begin{pmatrix}
			B^{(a)}_{k,k,6} & & & \\
			B^{(a)}_{k+1,k,4} & \ddots & & \\
			B^{(a)}_{k+2,k,2} & \ddots & \ddots & \\
			& \ddots & \ddots & B^{(a)}_{N-1,k,6} \\
			& & B^{(a)}_{N,k,2} & B^{(a)}_{N,k,4}
		\end{pmatrix}, \quad k = 0,1,2,\dots,N \\
	&\big( \text{so that } A^{(a)}_{y,0} \in \R^{(N+1)\times2N}, \quad A^{(a)}_{y,k} \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \big) \\
	C^{(a)}_{y,k} &:= 
		\begin{pmatrix}
			B^{(a)}_{k,k,1} & B^{(a)}_{k,k,3} & B^{(a)}_{k,k,5} & & \\
			& \ddots & \ddots & \ddots \\
			& & \ddots & \ddots & B^{(a)}_{N-1,k,5} \\
			& & & B^{(a)}_{N,k,1} & B^{(a)}_{N,k,3}
		\end{pmatrix}, \quad k = 1,2,\dots,N \\
	&\big( \text{so that } C^{(a)}_{y,1} \in \R^{2N\times(N+1)}, \quad C^{(a)}_{y,k} \in \R^{2(N-k+1)\times2(N-k+2)} \text{ for } k = 2,3,\dots,N \big)
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align}
	B^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			0 & \betaa_{n,k,0,j} \\
			\betaa_{n,k,1,j} & 0
		\end{pmatrix} \in \R^{2\times2}, (k \ne 1 \text { for } j \text{ odd}) \label{eqn:jacobisubblocksB1} \\
	B^{(a)}_{n,0,j} &:=
		\begin{pmatrix}
			0 & \betaa_{n,0,0,j}
		\end{pmatrix} \in \R^{1\times2}, j \text{ odd}  \label{eqn:jacobisubblocksB2} \\
	B^{(a)}_{n,1,j} &:=
		\begin{pmatrix}
			0 \\
			\betaa_{n,1,1,j}
		\end{pmatrix}\in \R^{2\times1}, j \text{ odd}  \label{eqn:jacobisubblocksB3} \\
\end{align}

For $J_z^{(a)}$, the sub-blocks $B^{(a)}_{z,k}$ have sub-block-bandwidths $(2,2)$, while the sub-blocks $A^{(a)}_{z,k}, \: C^{(a)}_{z,k}$ are zero matrices:
\begin{align*}
	A^{(a)}_{z,0} &= 0 \in \R^{(N+1)\times2N}, \quad A^{(a)}_{z,k} = 0 \in \R^{2(N-k+1)\times2(N-k)} \text{ for } k = 1,2,\dots,N \\
	C^{(a)}_{z,1} &= 0 \in \R^{2N\times(N+1)}, \quad C^{(a)}_{z,k} = 0 \in \R^{2(N-k+1)\times2(N-k+2)} \text{ for } k = 2,3,\dots,N \\
	B^{(a)}_{z,k} &:= 
		\begin{pmatrix}
			\Gamma^{(a)}_{k,k,2} & \Gamma^{(a)}_{k,k,3} & \\
			\Gamma^{(a)}_{k+1,k,1} & \ddots & \ddots & \\
			& \ddots & \ddots & \Gamma^{(a)}_{N-1,k,3} \\
			& & \Gamma^{(a)}_{N,k,1} & \Gamma^{(a)}_{N,k,2}
		\end{pmatrix}  \quad k = 0,1,2,\dots,N \\
	&\big( \text{so that } B^{(a)}_{z,0} \in \R^{(N+1)\times(N+1)}, \quad B^{(a)}_{z,k} \in \R^{2(N-k+1)\times2(N-k+1)} \text{ for } k = 1,2,\dots,N \big)
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align}
	\Gamma^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			\gamma_{n,k,j} & 0 \\
			0 & \gamma_{n,k,j}
		\end{pmatrix} \in \R^{2\times2}, \label{eqn:jacobisubblocksGamma1} \\
	\Gamma^{(a)}_{n,0,j} &:= \gammaa_{n,0,j}. \label{eqn:jacobisubblocksGamma2}
\end{align}

@@

In this section, we introduce new notation for a different ordering of the OP vector, in order to demonstrate how we can obtain a recurrence for the OPs in matrix-vector form, and derive the associated \enquote{Clenshaw matrices} -- we define the Clenshaw matrices as the matrices derived from the OP's $x,y,z$-multiplication relationships that form the multidimensional three-term recurrences. While in Section \ref{subsection:jacobimats} the ordering used is convenient for the application of Jacobi and differential operators to the vector of coefficients of a given function's expansion, the ordering we will use in this section is convenient for building the OPs and more importantly obtaining the \enquote{Clenshaw matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings.

@@

\begin{definition}\label{def:differentialoperators}
	Let $a$ be a nonnegative parameter, and $\tilde a \ge 2$ be a positive integer. Define the operator matrices $D_\varphi^{(a)}, \: W_\varphi^{(a)}, \: D_\theta, \: \calL^{(a)\to(a+\tilde a)}, \: \calL_W^{(a)\to(a-\tilde a)}, \: \Delta^{(1)}_W$ according to:
\begin{align*}
	\rho \pfpx{\varphi}{f} (x,y,z)&= \bigscop_N^{(a+1)}(x,y,z)^\top \: D_\varphi^{(a)} \: \mathbf{f}, \\
	\rho \ppx{\varphi}[\genjacw^{(a,0)}(z) \: f(x,y,z)] &= \bigW_N^{(a-1)}(x,y)^\top \: W_\varphi^{(a)}\: \mathbf{f}, \\
	\pfpx{\theta}{f}(x,y,z) &= \bigscopNa(x,y,z)^\top \: D_\theta \: \mathbf{f}, \\
	\DeltaS f(x,y,z) &= \bigscopN^{(a+\tilde a)}(x,y,z)^\top \: \calL^{(a)\to(a+\tilde a)} \: \mathbf{f}, \\
	\DeltaS \big( \genjacw^{(a,0)}(z) \: f(x,y,z) \big) &= \bigW_N^{(a-\tilde a)}(x,y,z)^\top \: \calL_W^{(a)\to(a-\tilde a)} \: \mathbf{f},  \quad \text{ (for } a \ge 2 \text{ only)} \\
	\DeltaS \big( \genjacw^{(1,0)}(z) \: f(x,y,z) \big) &= \bigscopN^{(1)}(x,y,z)^\top \: \Delta^{(1)}_W \: \mathbf{f}, \quad \text{ (for } a = 1 \text{ only)}
\end{align*}
\end{definition}

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\begin{align*}
	u(x,y,z) &= 0 \quad \text{for } (x,y,z) \in \Omega, \\
	\pfpx{n}{u}(x,y,z) &= \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
\end{align*}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(\vec{x}) = {\vec{x} \over \norm{\vec{x}}}$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\begin{align*}
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
\end{align*}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in Figure \ref{fig:sparsity}.
